# Recoding Variables

```{r, eval=TRUE, echo=FALSE}
library(tidyverse)
load("our_gss.rda")
```

Today's venture concerns univariate analysis, i.e. the quantitative description of a single variable. Before we do that, however, we need to familiarize ourselves with some data-cleaning procedures. 

Recoding a variable involves manipulating the underlying structure of our variable such that we can use it for analysis. We did a little recoding during the last unit when we converted character vectors into factor variables. This allowed us to align R data types with the appropriate level of measurement.

There are also occasions when we need a variable to be translated from one level of measurement to another. For example, we may want to convert a ratio variable for "number of years of education" into an ordinal variable reflecting categories like "less than high school", "high school diploma", "Associates degree", and so on. 

We may also want to collapse the categories of ordinal variables for some analyses. Take a variable like `partyid`, which has the following categories:
```{r, eval=TRUE, echo=FALSE}
our_gss |> count(partyid)
```



